\vspace{-1.5em} \bf A general and efficient representation of ancestral recombination graphs

### 2023-10-17 paper.tex
Add SCAR paper
Closes #392

\linenumbers
\section{Introduction}
\label{sec-intro}
Estimating the genetic genealogy of a set of genome sequences
under the influence of recombination,
usually known as an Ancestral Recombination Graph (ARG), is a long-standing
goal in genetics.
Broadly speaking, an ARG describes the different paths of genetic inheritance
caused by recombination, and encodes a sequence of correlated genealogical
trees along the genome.
Recent breakthroughs
in large-scale inference
methods~\citep{rasmussen2014genome,kelleher2019inferring,speidel2019method,
schaefer2021ancestral,wohns2022unified,zhang2023biobank,zhan2023towards}
and data representation~\citep{kelleher2016efficient,kelleher2018efficient}
have raised the realistic prospect of ARG-based analysis becoming a standard part
of the population and statistical genetics toolkit~\citep{hejase2020summary}.
Applications using inferred ARGs as input have begun to
appear~\citep{osmond2021estimating,fan2022genealogical,hejase2022deep,zhang2023biobank,
nowbandegani2023extremely,ignatieva2023distribution,fan2023likelihood}
and many more are sure to
follow~\citep{harris2019database,harris2023using}.
See \citet{lewanski2023era} for a biologically oriented introduction to ARGs.

@@

\section{Computational efficiency and data interchange}
\label{sec-efficiency}
ARG inference methods have been successfully applied to vast datasets,
including 500,000 genotyped humans~\citep{kelleher2019inferring,zhang2023biobank}
from UK Biobank~\citep{bycroft2018genome}
and over a million SARS-Cov-2 whole genomes~\citep{zhan2023towards}.
Even larger datasets are
available~\citep[e.g.][]{halldorsson2022sequences} % TODO more
and it seems inevitable that ARGs will be inferred for them.
The efficiency with which we can store and analyse these inferred
ARGs is therefore of great importance, and a key factor
in the ultimate success of downstream ARG-based methods.
Another key factor in how important ARG-based methods become
across the wide range of possible applications
% TODO more?
areas~\citep{rasmussen2014genome,harris2019database,
hejase2020summary,schaefer2021ancestral,
fan2022genealogical,wohns2022unified,harris2023using,
nowbandegani2023extremely,ignatieva2023distribution}
is how well inference methods and downstream analysis applications
interoperate. Without a well-defined and efficient shared format
for ARG data interchange,
users will suffer from problems with converting data from one program's
output to another, an unfortunate hallmark of population genetics
software for many years~\citep{excoffier2006computer}.
While current methods tend to tightly couple downstream analysis
of the inferred ARG with the inference
itself within the same software package,
this approach scales poorly in terms of software development
effort, and is ultimately not compatible with the widespread use
of ARGs for routine data analysis, and a healthy and diverse
software ecosystem.

The gARG encoding discussed and concretely defined in this manuscript
leads to highly efficient storage and processing of ARG data,
and has already been in use for several years.
The succinct tree sequence data structure (usually known as a ``tree sequence''
for brevity)
is a concrete encoding of a gARG, as discussed in this article.
It was originally developed as part of the \texttt{msprime}
simulator~\citep{kelleher2016efficient} and has subsequently been
extended and applied to forward-time
simulations~\citep{kelleher2018efficient,haller2018tree},
inference from data~\citep{kelleher2019inferring,wohns2022unified},
and calculation of population genetics statistics~\citep{ralph2020efficiently}.
The succinct tree sequence encoding extends the basic definition
of a gARG provided here by stipulating a
simple tabular representation and nodes and edges,
and also defining a concise and lossless representation of
sequence variation using the ``site'' and  ``mutation'' tables.
(Technically, edges in tree sequence terminology would be better
described as ``edge-intervals'', as each describes a single contiguous
interval of genome inheritance between a pair of nodes. This
denormalisation of the gARG data model in \texttt{tskit} is for efficiency purposes.)
The \texttt{tskit} library is a liberally
licensed open source toolkit that provides a comprehensive suite
of tools for working with gARGs (encoded as a succinct tree sequence).
Based on core functionality written
in C, it provides interfaces in C, Python and Rust.
Tskit is mature software, widely used in population genetics, and
has been incorporated into several downstream
applications~\citep[e.g.,][]{haller2019slim,speidel2019method,
adrion2020community,
terasaki2021geonomics,
baumdicker2021efficient,
fan2022genealogical,korfmann2022weak,
mahmoudi2022bayesian,petr2022slendr,rasmussen2022espalier,
zhang2023biobank,nowbandegani2023extremely,ignatieva2023distribution,
fan2023likelihood}.

@@

% % Do downstream applications actually make use of such full ARGs?
Besides the inherent limitations that exist on inferring fully
precise ARGs from data,
we should also consider the value that such precise estimates provide
for downstream applications.
% To date, few make use such detailed recombination information.
Most applications work by examining local trees independently.
For example, the \relate\ selection test~\citep{speidel2019method}
obtains $p$-values by computing clade size probabilities conditional
on the timing of coalescence events in a given local tree.
In their method
for estimating dispersal rates and the locations of genetic
ancestors,
\cite{osmond2021estimating} downsample trees along the genome
so that they can be regarded as approximately independent.
Similarly, \cite{fan2023likelihood} compute the likelihood
of an ARG under a particular demographic model as the product
over a sample of widely-separated local trees, assumed to be independent.
The SIA method for detecting selection~\citep{hejase2022deep}
encodes local trees as a set of lineage counts at discrete
time intervals, and uses these as feature for a
type of machine learning algorithm
that takes ``temporal'' correlations into account. Thus,
while SIA does use some of the information about local tree correlation,
clearly much of the detail about recombination
events in an ARG is lost.
The \texttt{tsdate} algorithm (and related naive estimator
of ancestral location) uses much more between-tree
information~\citep{wohns2022unified}, explicitly using the gARG
encoding to reason about ancestral nodes.
The main application for fully precise ARGs thus far has been
to compute a likelihood under the coalescent with
recombination~\citep[e.g.][]{kuhner2000maximum,mahmoudi2022bayesian},
which currently requires the details of all recombination
events to be known.

@@

% Ordered alphabetically by surname
Gregor Gorjanc acknowledges support from the BBSRC ISP grant to The Roslin Institute (BBS/E/D/30002275, BBS/E/RL/230001A, and BBS/E/RL/230001C).
% Anastasia Ignatieva acknowledges support from TODO
Jerome Kelleher acknowledges support from the Robertson Foundation.
Jere Koskela acknowledges support from EPSRC research grant EP/V049208/1.
% Anthony Wilder Wohns acknowledges support from TODO

>>>>

\linenumbers
\section{Introduction}
\label{sec-intro}
Estimating the genetic genealogy of a set of genome sequences
under the influence of recombination,
usually known as an Ancestral Recombination Graph (ARG), is a long-standing
goal in genetics.
Broadly speaking, an ARG describes the different paths of genetic inheritance
caused by recombination, and encodes a sequence of correlated genealogical
trees along the genome.
Recent breakthroughs
in large-scale inference
methods~\citep{rasmussen2014genome,kelleher2019inferring,speidel2019method,
schaefer2021ancestral,wohns2022unified,zhang2023biobank,zhan2023towards}
% and data representation~\citep{kelleher2016efficient,kelleher2018efficient}
have raised the realistic prospect of ARG-based analysis becoming a standard part
of the population and statistical genetics toolkit~\citep{hejase2020summary}.
Applications using inferred ARGs as input have begun to
appear~\citep{osmond2021estimating,
fan2022genealogical,
hejase2022deep,
guo2022recombination,
zhang2023biobank,
nowbandegani2023extremely,
ignatieva2023distribution,
fan2023likelihood}
and many more are sure to
follow~\citep{harris2019database,harris2023using}.
% TODO finish this sentence with "and more reasons why ARGs are cool"
See \citet{lewanski2023era} for a biologically oriented introduction to ARGs.

@@

\section{Computational efficiency and data interchange}
\label{sec-efficiency}
ARG inference methods have been successfully applied to vast datasets,
including 500,000 genotyped humans~\citep{kelleher2019inferring,zhang2023biobank}
from UK Biobank~\citep{bycroft2018genome}
and over a million SARS-Cov-2 whole genomes~\citep{zhan2023towards}.
Even larger datasets are
available~\citep[e.g.][]{halldorsson2022sequences} % TODO more
and it seems inevitable that ARGs will be inferred for them.
The efficiency with which we can store and analyse these inferred
ARGs is therefore of great importance, and a key factor
in the ultimate success of downstream ARG-based methods.
Another key factor in how important ARG-based methods become
across the wide range of possible applications
% TODO get rid of this, it's a duplication of the introduction
areas~\citep{rasmussen2014genome,harris2019database,
hejase2020summary,schaefer2021ancestral,
fan2022genealogical,wohns2022unified,harris2023using,
nowbandegani2023extremely,ignatieva2023distribution}
is how well inference methods and downstream analysis applications
interoperate. Without a well-defined and efficient shared format
for ARG data interchange,
users will suffer from problems with converting data from one program's
output to another, an unfortunate hallmark of population genetics
software for many years~\citep{excoffier2006computer}.
While current methods tend to tightly couple downstream analysis
of the inferred ARG with the inference
itself within the same software package,
this approach scales poorly in terms of software development
effort, and is ultimately not compatible with the widespread use
of ARGs for routine data analysis, and a healthy and diverse
software ecosystem.

The gARG encoding discussed and concretely defined in this manuscript
leads to highly efficient storage and processing of ARG data,
and has already been in use for several years.
The succinct tree sequence data structure (usually known as a ``tree sequence''
for brevity)
is a concrete encoding of a gARG, as discussed in this article.
It was originally developed as part of the \texttt{msprime}
simulator~\citep{kelleher2016efficient} and has subsequently been
extended and applied to forward-time
simulations~\citep{kelleher2018efficient,haller2018tree},
inference from data~\citep{kelleher2019inferring,wohns2022unified},
and calculation of population genetics statistics~\citep{ralph2020efficiently}.
The succinct tree sequence encoding extends the basic definition
of a gARG provided here by stipulating a
simple tabular representation and nodes and edges,
and also defining a concise and lossless representation of
sequence variation using the ``site'' and  ``mutation'' tables.
(Technically, edges in tree sequence terminology would be better
described as ``edge-intervals'', as each describes a single contiguous
interval of genome inheritance between a pair of nodes. This
denormalisation of the gARG data model in \texttt{tskit} is for efficiency purposes.)
The \texttt{tskit} library is a liberally
licensed open source toolkit that provides a comprehensive suite
of tools for working with gARGs (encoded as a succinct tree sequence).
Based on core functionality written
in C, it provides interfaces in C, Python and Rust.
Tskit is mature software, widely used in population genetics, and
has been incorporated into several downstream
applications~\citep[e.g.,][]{haller2019slim,speidel2019method,
adrion2020community,
terasaki2021geonomics,
baumdicker2021efficient,
fan2022genealogical,
guo2022recombination,
korfmann2022weak,
mahmoudi2022bayesian,
petr2022slendr,
rasmussen2022espalier,
zhang2023biobank,
nowbandegani2023extremely,
ignatieva2023distribution,
fan2023likelihood}.

@@

% % Do downstream applications actually make use of such full ARGs?
Besides the inherent limitations that exist on inferring fully
precise ARGs from data,
we should also consider the value that such precise estimates provide
for downstream applications.
% To date, few make use such detailed recombination information.
Most applications work by examining local trees independently.
For example, the \relate\ selection test~\citep{speidel2019method}
obtains $p$-values by computing clade size probabilities conditional
on the timing of coalescence events in a given local tree.
In their method
for estimating dispersal rates and the locations of genetic
ancestors,
\cite{osmond2021estimating} downsample trees along the genome
so that they can be regarded as approximately independent.
Similarly, \cite{fan2023likelihood} compute the likelihood
of an ARG under a particular demographic model as the product
over a sample of widely-separated local trees, assumed to be independent.
The SIA method for detecting selection~\citep{hejase2022deep}
encodes local trees as a set of lineage counts at discrete
time intervals, and uses these as feature for a
type of machine learning algorithm
that takes ``temporal'' correlations into account. Thus,
while SIA does use some of the information about local tree correlation,
clearly much of the detail about recombination
events in an ARG is lost.
The \texttt{tsdate} algorithm (and related naive estimator
of ancestral location) uses much more between-tree
information~\citep{wohns2022unified}, explicitly using the gARG
encoding to reason about ancestral nodes.
The main application for fully precise ARGs thus far has been
to compute a likelihood under the coalescent with
recombination~\citep[e.g.][]{kuhner2000maximum,mahmoudi2022bayesian,
guo2022recombination},
which currently requires the details of all recombination
events to be known.

@@

% Ordered alphabetically by surname
Gregor Gorjanc acknowledges support from the BBSRC ISP grant to The Roslin Institute
(BBS/E/D/30002275, BBS/E/RL/230001A, and BBS/E/RL/230001C).
% Anastasia Ignatieva acknowledges support from TODO
% TODO Jerome: add EPSRC and NIH grants.
Jerome Kelleher acknowledges support from the Robertson Foundation.
Jere Koskela acknowledges support from EPSRC research grant EP/V049208/1.
% Anthony Wilder Wohns acknowledges support from TODO
